Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-05-05 application.tex
slight revision of sec 7, shortened sec 8 (moved viscosity disscussion here), shortened sec 6.4 (vert non-iso)

As an application of our linear models, we estimate where in a
PPD do the thermodynamic conditions allow the VSI to
operate. We first model the cooling time $t_c$ based on realistic
disk properties, which will yield a non-uniform dimensionless cooling
time $\beta(r,z;\khat)$. We then compare this to the critical cooling time
$\beta_\mathrm{crit}$ derived above. Finally, we explicitly compute 
linear growth rates numerically in order to characterize 
growth times expected in PPDs.

@@

For perturbations with lengthscales $l\gg l_\mathrm{rad}$, we assume
temperature fluctuations are smoothed out by radiative diffusion and
model this by writing the linearized cooling function as

@@

Since the VSI is characterized by vertically-elongated,
radially-narrow disturbances, we simplify Eq. \ref{diff_cool_proper} 
by retaining only radial derivatives of the perturbations, so that 
\begin{align}\label{diff_cool_approx}
  % \frac{P}{\rho T C_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
  %  T\right) &
  \delta\Lambda \to %-\frac{k_\mathrm{rad}}{\rho
  % C_v}k_x^2P\left(\frac{\delta T}{T}\right)%\notag\\
  % \equiv 
  %-\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right) =
  -\eta k_x^2 P \frac{\delta T}{T}, %=
  %-\frac{P}{t_\mathrm{diff}}\frac{\delta T}{T},
  %\frac{P}{t_\mathrm{diff}}\left(\frac{delta T}{T}\right), 
\end{align}
\citepalias[see also][]{barker15} where $\eta=k_\mathrm{rad}/\rho C_v$ is
the diffusion coefficient and 
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 C_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. From our
definition of thermal relaxation (Eq. \ref{thermal_relax}), 
we thus identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\OmK^{-1}}{\hat{\eta}\khat^2}, 
\end{align}

@@

In Fig. \ref{mmsn_bcrit_bcool}, we compare $\beta$ to
$\beta_\mathrm{crit}$ for $r_\mathrm{AU}\in[1,100]$, as well as for different
opacity scales. Notice that $\beta$ increases for either large or
small radii. For sufficiently large $r$, disturbances are generally
in the optically-thin cooling regime, characterized graphically by the
over-lapping curves. 
%, where $\beta$ is independent of
%$z$ and the perturbation lengthscale $\khat$.
In this case, we may
directly compare $\beta$ to $\beta_\mathrm{crit}$ and infer that the
VSI can operate at 10s of AU in the fiducial MMSN disk (with
$\hat{\kappa}_d=1$).  However, at small radii, $\beta$ depends on both
$z$ and $\khat$, as the disturbance enters the optically-thick limit
at the midplane. This regime is characterized by the distinct curves for each
$\khat$ toward decreasing $r$.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$ and small $r$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane at, say $r_\mathrm{AU}\sim 1$, the VSI would only occur on very small
radial lengthscales $\ll O(10^{-2}H)$. It is, however, unclear if such
small-scale disturbances are dynamically important.

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the fundamental mode with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. In fact, we either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$. We discard
surface modes for this discussion since they are not well-defined 
in vertically isothermal disk models \citepalias{barker15} and may be 
subject to viscous decay in real disks (see \S\ref{caveats_visc}).

@@

The fact that we can obtain the VSI in the despite $\beta >
\beta_\mathrm{crit}$ near the midplane (e.g. $\khat = 30$ at
$r_\mathrm{AU} =5$, see Fig. \ref{beta_compare}) suggests  that the instability
can operate provided $\beta < \beta_\mathrm{crit}$ for a sufficiently large
fraction of the vertical domain. However, the normalized growth times
become much longer. For example, for $\khat=30$ we find
$t_\mathrm{grow}\sim 40$ orbits at $r_\mathrm{AU}=50$, but $t_\mathrm{grow}\sim
400$ orbits at $r_\mathrm{AU}=5$ (and rapidly increases to $> 10^3$ orbits for
smaller $r$). Thus, introducing an optically-thick midplane 
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the VSI in 
    the MMSN. These are calculated using the thermal relaxation
    function given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

>>>>

As an application of our linear models, we estimate where in a
PPD do the thermodynamic conditions allow the VSI to
operate. We first model the cooling time $t_c$ based on realistic
disk properties, which yields a non-uniform dimensionless cooling
time $\beta(r,z;\khat)$. We then compare this to the critical cooling time
$\beta_\mathrm{crit}$ derived above. Finally, we explicitly compute 
linear growth rates numerically, using $\beta(r,z;\khat)$, to
characterize growth times expected in a typical PPD.

@@

For perturbations with lengthscales $l\gg l_\mathrm{rad}$, we assume
temperature fluctuations are smoothed out by radiative diffusion. We 
model this by writing the linearized cooling function as

@@

Since the VSI is characterized by vertically-elongated,
radially-narrow disturbances, we simplify Eq. \ref{diff_cool_proper} 
by retaining only radial derivatives of the perturbations, so that 
\begin{align}\label{diff_cool_approx}
  % \frac{P}{\rho T C_v} \nabla\cdot\left(k_\mathrm{rad}\nabla\delta
  %  T\right) &
  \delta\Lambda \to %-\frac{k_\mathrm{rad}}{\rho
  % C_v}k_x^2P\left(\frac{\delta T}{T}\right)%\notag\\
  % \equiv 
  %-\eta k_x^2 \left(\delta P - \frac{P}{\rho}\delta\rho\right) =
  -\eta k_x^2 P \frac{\delta T}{T}, %=
  %-\frac{P}{t_\mathrm{diff}}\frac{\delta T}{T},
  %\frac{P}{t_\mathrm{diff}}\left(\frac{delta T}{T}\right), 
\end{align}
\citepalias[see also][]{barker15} where $\eta=k_\mathrm{rad}/\rho C_v$ is
the diffusion coefficient and 
is given in terms of physical disk parameters as 
\begin{align}\label{eta_def}
  \eta = \frac{16\sigma_s T^3}{3\kappa_d\rho^2 C_v}, 
\end{align}
where $\sigma_s$ is the Stefan-Boltzmann constant. From our
definition of thermal relaxation (Eq. \ref{thermal_relax}), 
we identify the scale-dependent thermal relaxation
time for diffusion as 
\begin{align}\label{tc_diff_cool} 
  t_\mathrm{diff} = \frac{1}{\eta k_x^2}.%\equiv \frac{\OmK^{-1}}{\hat{\eta}\khat^2}, 
\end{align}

@@

In Fig. \ref{mmsn_bcrit_bcool}, we compare $\beta$ to
$\beta_\mathrm{crit}$ for $r_\mathrm{AU}\in[1,100]$, as well as for different
opacity scales. Notice that $\beta$ increases for either large or
small radii. For sufficiently large $r$, disturbances are generally
in the optically-thin cooling regime, characterized graphically by the
over-lapping curves. 
%, where $\beta$ is independent of
%$z$ and the perturbation lengthscale $\khat$.
In this case, we may use the criterion $\beta<\beta_\mathrm{crit}$ to
infer that the VSI can operate at 10s of AU in the fiducial MMSN disk (with
$\hat{\kappa}_d=1$).  However, at small radii, $\beta$ depends on both
$z$ and $\khat$, as the disturbance enters the optically-thick limit
at the midplane. This regime is characterized by the distinct curves for each
$\khat$ toward decreasing $r$.

@@

We remark that it is possible to satisfy $\beta < \beta_\mathrm{crit}$
for sufficiently large $\khat$ and small $r$. We illustrate this in 
Fig. \ref{mmsn_bcrit_bcool_mink} by plotting at each radius the value
of $\khat$ such that $\beta = \beta_\mathrm{crit}$. If we consider the
midplane at, say $r_\mathrm{AU}\sim 1$, the VSI would only occur on 
radial lengthscales $\ll O(10^{-2}H)$. It is, however, unclear if such
small-scale disturbances are dynamically important.

@@

In Fig. \ref{mmsn_overall} we plot characteristic growth timescales
$t_\mathrm{grow} \equiv 1/\sigma$ of the VSI with a range of
wavenumbers, in units of the local orbital time $P_\mathrm{orb}\equiv
2\pi/\OmK$. We either only find the fundamental mode, or 
that it is the most unstable, except for the appearance of some 
surface modes for large wavenumbers $\khat\gtrsim 50$. We discard
surface modes for this discussion since they are not well-defined 
in vertically isothermal disks with an imposed boundary,
\citepalias{barker15} and may be subject to viscous decay in reality
(see \S\ref{caveats_visc}).

@@

The fact that we can obtain the VSI at small radii, despite $\beta >
\beta_\mathrm{crit}$ near the midplane (e.g. $\khat = 30$ at
$r_\mathrm{AU} =5$, see Fig. \ref{beta_compare}), suggests  that the
instability can operate provided $\beta < \beta_\mathrm{crit}$ for a
sufficiently large fraction of the vertical domain. However, the
normalized growth times become much longer. For example, for
$\khat=30$ we find $t_\mathrm{grow}\sim 40$ orbits at
$r_\mathrm{AU}=50$, but $t_\mathrm{grow}\sim 400$ orbits at
$r_\mathrm{AU}=5$ (and rapidly increases to $> 10^3$ orbits for
smaller $r$). Thus, introducing an optically-thick midplane
significantly reduces the efficiency of the VSI despite $\beta <
\beta_\mathrm{crit}$ away from the midplane.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/eigen_compare_grow.ps}
  \caption{Characteristic growth times of the VSI in 
    the fiducial MMSN. These are calculated using the thermal relaxation
    function given by Eq. \ref{beta_mmsn_simp}. 
    \label{mmsn_overall}}    
\end{figure}

### 2015-05-05 summary.tex
rough re-write of sec 7 done by mkl
slight revision of sec 7, shortened sec 8 (moved viscosity disscussion here), shortened sec 6.4 (vert non-iso)

\section{Summary and discussion}\label{summary}
In this paper we have studied the axisymmetric vertical shear 
instability (VSI) in astrophysical disks that arise from the 
height dependence of the equilibrium angular velocity profile 
$\Omega=\Omega(r,z)$ in a baroclinic disk. Vertical shear is generally
expected in astrophysical disks due to baroclinicity. This may arise 
from a radial variation in the disk temperature, or more generally in
the disk entropy. However, astrophysical disks are also typically
sub-adiabatically stratified for stability against vertical
convection. In this case, the VSI requires a short thermal or cooling
timescale $t_c$ in order to operate.

Physically, a short thermal timescale is needed to reduce the  
stabilizing effect of buoyancy forces in a stably stratified
disk. Because vertical shear is weak in thin disks, $t_c\Omega\ll 1$
is generally needed for the VSI to operate, otherwise buoyancy would
be strongly stabilizing. Our primary goal was to quantify this
requirement, thereby provide a simple way to assess 
whether or not the VSI is relevant in realistic disks.

To do so, we generalized previous vertically global linear 
calculations of the VSI \citepalias{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with 
a source term in that relaxes the disk
temperature to its initial value on a timescale
$t_c$. % This `thermal relaxation' model allows
% the thermodynamic response of the disk to be controlled by varying
% $t_c$, and has been employed in previous numerical simulations of the
% VSI \citepalias{nelson13}.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. 
We mostly focused on vertically isothermal disks as 
this is the relevant case for protoplanetary disks (PPDs) passively
irradiated by its central star \citep{chiang97}. It also permits a
simplified analytical discussion by extending the
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) thermal relaxation. We obtained explicit solutions 
describing the VSI with thermal relaxation in the thin disk limit.

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{ h |q|}{\gamma-1}\OmK^{-1}, 
\end{align*}
below which the VSI operates effectively. While $t_\mathrm{crit}$ was
derived for the lowest order VSI mode --- the `fundamental  
corrugation mode' --- we find from numerical calculations that
$t_\mathrm{crit}$ can be used to characterize the general
stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have therefore provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O( h)$, which is
quite short for thin disks ($ h\ll 1$), because vertical shear is weak.

We applied our linear models to assess the applicability of the VSI in
PPDs. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $t_c$ based on realistic
disk properties and dust opacity. 
%This results in $t_c$ being a
% function of position and the perturbation lengthscale of interest. 
We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. It is possible to have the VSI at smaller
radii, but only on radial lengthscales signficantly shorter
than the disk scaleheight. However, such small-scale perturbations may 
be subject to viscous decay in a real disk.

@@

\subsubsection{Viscosity} 
% Depending on where the VSI operates in a
% disk, additional effects may become important. For example, in the
% outer parts of a PPD, where we estimate the VSI 
% operates, disk self-gravity (SG) may play a role. Vertical
% SG may favor the VSI by increasing $|\p_z\Omega|$ through
% enhancing the density stratification 
% (Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

We have neglected viscosity. \citetalias{nelson13} find that a
dimensionless viscosity of $\alpha\lesssim O(10^{-4})$ is required for the
VSI; while \citetalias{barker15} find that inclusion of viscosity is needed for 
convergence when studying surface modes in radially global disk
models. %but not body modes? 
Viscosity may also affect the
thermodynamic requirement for the VSI, because viscous heating
needs to be self-consistently included in the energy equation. This
should be investigated in future works.

However, disturbances with large $\khat$ may be subject to viscous
decay in a real disk. Adopting the standard precription for the
kinematic viscosity $\nu = \alpha c_s H$ \citep{shakura73}, where
$\alpha$ is the dimensionless viscosity parameter, 
the viscous timescale for a perturbation lengthscale 
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These low viscosity levels are required for the
VSI, see][]{nelson13}.

@@

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects can be represented by a thermal
relaxation timescale $t_c$. For  
application to PPDs we connected $t_c$ to the disk
properties such as surface density, temperature and opacity, making
$t_c$ depend on position and the perturbation 
lengthscales. One might improve the realism further by generalizing
the opacity model adopted here to be density (and hence height)
dependent.  
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. % The linear problem then
% involves a Laplacian operator and additional boundary
% conditions must be supplemented. 
We will consider these improvements in a
follow-up study. %additional heat source

>>>>

\section{Summary and discussion}\label{summary}
In this paper we have studied the axisymmetric vertical shear 
instability (VSI) in astrophysical disks that arise from the 
height dependence of the equilibrium angular velocity profile,  
$\p_z\Omega\neq 0$. Vertical shear is generally
expected in astrophysical disks due to baroclinicity ($\nabla
P\times\nabla \rho \neq 0$). This may be due
to a radial variation in the disk temperature, or more generally in
the disk entropy. However, astrophysical disks are also typically
sub-adiabatically stratified for stability against vertical
convection ($N_z^2\geq 0$). In this case, the VSI requires a short
thermal or cooling timescale $t_c$ in order to operate.

Physically, a short thermal timescale is needed to reduce the  
stabilizing effect of buoyancy forces. Because vertical shear is weak
in thin disks, very rapid cooling, corresponding to $t_c\Omega\ll 1$, 
is generally needed for the VSI, otherwise buoyancy is strongly
stabilizing. Our primary goal was to quantify this requirement,
thereby provide a simple way to assess whether or not the VSI is
relevant in realistic disks.

To do so, we generalized previous vertically global, radially local linear 
calculations of the VSI \citepalias{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with 
a source term in that relaxes the disk 
temperature to its initial value on a timescale
$t_c$. 
% This `thermal relaxation' model allows
% the thermodynamic response of the disk to be controlled by varying
% $t_c$, and has been employed in previous numerical simulations of the
% VSI \citepalias{nelson13}.  
%In accordance with the above studies, which show that the VSI occurs for 
%perturbations with small radial lengthscales, we applied the 
%radially local approximation in the linearized fluid equations. 
We mostly focused on vertically isothermal disks as this is the
relevant case for protoplanetary disks (PPDs) passively 
irradiated by its central star \citep{chiang97}. It also permits a
simplified analytical discussion by extending the 
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) cooling. We obtained explicit solutions 
describing the VSI with thermal relaxation in the thin disk limit.

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{ h |q|}{\gamma-1}\OmK^{-1}, 
\end{align*}
below which the VSI operates effectively. 
%While $t_\mathrm{crit}$ was
%derived for the lowest order VSI mode --- the `fundamental  
%corrugation mode' --- we find from numerical calculations that
%$t_\mathrm{crit}$ can be used to characterize the general
%stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have thus provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O( h)$, which is 
quite short for thin disks ($ h\ll 1$), because vertical shear is weak.

We applied our linear models to assess the applicability of the VSI in
PPDs. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $t_c$ based on realistic
disk properties and dust opacity. 
%This results in $t_c$ being a
% function of position and the perturbation lengthscale of interest. 
We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. It is possible to have the VSI at smaller
radii, but only on radial lengthscales signficantly shorter
than the disk scaleheight. %However, such small-scale perturbations may 
%be subject to viscous decay in a real disk.

@@

\subsubsection{Viscosity}\label{caveats_visc}
% Depending on where the VSI operates in a
% disk, additional effects may become important. For example, in the
% outer parts of a PPD, where we estimate the VSI 
% operates, disk self-gravity (SG) may play a role. Vertical
% SG may favor the VSI by increasing $|\p_z\Omega|$ through
% enhancing the density stratification 
% (Eq. \ref{vertical_shear}). %sg in pert? no because kx>>1. but maybe for kx~1

We have neglected viscosity. 
% while \citetalias{barker15} find that inclusion of viscosity is needed for 
% convergence when studying surface modes in radially global disk
% models. %but not body modes? 
% Viscosity may also affect the
% thermodynamic requirement for the VSI, because viscous heating
% needs to be self-consistently included in the energy equation. This
% should be investigated in future works.   
An important effect of viscosity is setting a minimum wavelength (or 
maximum wavenumber) for which the VSI is relevant. Disturbances with
large $\khat\equiv k_xH$ may be subject to viscous decay. Adopting the standard
precription for the kinematic viscosity $\nu = \alpha c_s H$
\citep{shakura73}, where $\alpha$ is the dimensionless viscosity
parameter, the viscous timescale for a perturbation lengthscale  
$l\sim 1/k_x$ is $t_\mathrm{visc} = 1/\alpha\khat^2\Omega $. 
Setting $t_\mathrm{visc}$ to be longer than the
minimum VSI growth timescale $1/h\Omega$ implies 
$\khat^2\lesssim h/\alpha$. Inserting $h \sim 0.05$,
we find only modes with $\khat\lesssim 20$ are relevant for
$\alpha\sim 10^{-4}$ or $\khat\lesssim 70$ for $\alpha\sim
10^{-5}$. \citepalias[These small viscosity values are required for
the VSI to operate, see][]{nelson13}.

%\citetalias{nelson13} find that a
%dimensionless viscosity of $\alpha\lesssim O(10^{-4})$ is required for the
%VSI.

The above argument justifies our focus on moderate
wavenumbers, $\khat=O(10)$, which precludes extremely small-scale
disturbances such as surface modes. Nevertheless, future calculations
should explicitly account for viscosity. In particular, viscous
heating may influence the thermodynamic dependence of the VSI.

% needs to be self-consistently included in the energy equation.

@@

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects can be represented by a thermal
relaxation timescale $t_c$. For  
application to PPDs we connected $t_c$ to the disk
properties such as surface density, temperature and opacity, making
$t_c$ depend on position and the perturbation 
lengthscales. One might improve the realism further by generalizing
the opacity model adopted here to be density (and hence height)
dependent.  
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. 
% done roughly in this work -
% a proper study 
% involves a Laplacian operator and additional boundary
% conditions must be supplemented. 
We will consider these improvements in a
follow-up study. %additional heat source
